The role of cognateness in non-native spoken word recognition

### 2024-05-03 manuscript/manuscript.qmd
First results

library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
library(gtExtras)
library(brms)
library(tidybayes)

@@

# import objects
tar_config_set(store = here::here("_targets"),
                        script = here::here("_targets.R"))
tar_load_globals()
tar_load(participants)
tar_load(exp_participants)
tar_load(quest_participants)
tar_load(stimuli)

theme_set(theme_ggdist())

@@

Listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. However, there are some sources of information which may make non-native speech comprehension easier. One of the sources of this increased difficulty stems from the possible mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener’s native language. For example, imagine a Spanish native, with no previous familiarity with any other language, who listens for the first time to French may encounter the word /`r knitr::asis_output("\U0070\U0254\U0281\U0074")`/ (*porte*), translation of door in French. The voiced uvular fricative consonant /`r knitr::asis_output("\U0281")`/ and the open-mid back rounded vowel /`r knitr::asis_output("\U0254")`/ do not exist in Spanish as phonemic categories. These discrepancies can nonetheless be perceived as allophonic variations the native (Spanish) phonemes /r/ and /o/ [@best2001discrimination], and the non-native speech signal may engage lexical processing mechanisms [@weber2004lexical], leading to the activation and selection, and therefore the recognition of its translation to Spanish /`r knitr::asis_output("\U02C8\U0070\U0077\U0065\U0072\U002E\U0074\U0061")`/ (*puerta*). The recognition of this French, unfamiliar word relies almost entirely on its similarity with its Spanish translation. Conversely, imagine now that the same Spanish listener hears English for the first time and encounters the word /`r knitr::asis_output("\U0064\U0254\U02D0")`/ (*door*). This time, phonological similarity is of little help: /`r knitr::asis_output("\U02C8\U0070\U0077\U0065\U0072\U002E\U0074\U0061")`/ and /`r knitr::asis_output("\U0064\U0254\U02D0")`/ share very few phonemes. It would come as a surprise if the listener was able to translate the English word successfully, if not by relying on previously learnt word-concept associations (which they lack, as they are unfamiliar to English).

@@

Early theoretical accounts of this 'Cognate effect' proposed that the impact of cognateness on lexical processing depends on bilinguals’ proficiency in their second language (L2) [@potter1984lexical; see @chen1989patterns for similar results]. The reasoning behind this claim was the assumption that second-language learners start acquiring words in their L2 by associating them to their translation in their L1 (lexical route), instead of directly to their shared concept (conceptual route). This word-to-word connectivity should be sensitive to form-similarity (i.e., cognateness): the more similar the two word- forms are, the stronger the connection. As learners become more proficient, the connection between L2 representations and their meaning grows stronger, and L2 word processing becomes less reliant on word-word connections between L1 and L2 representations. Cognates subsequently exert less impact on lexical processing in proficient speakers [see @andras2022cognate for recent experimental evidence for this claim]. The Revised Hierarchical Model [RHM, @kroll1994category] captured this hypothesis and predicted that translating words from L1 to L2 should take longer than translating from L2 to L1. The rationale behind this prediction was that L1-to-L2 translation relies more strongly on word-to-word links between L1 and L2 representations, whereas L2-to-L1 translation relies more strongly on the mediation between the concept and the two word- forms. L1-to-L2 translation should therefore be more sensitive to the form similarity between the L1 and the L2 representations: cognate words should be retrieved faster than non-cognate words during L1-to-L2 translation.

@@

These findings support a soft version of the RHM model, in which both the conceptual and the lexical translation routes are active during translation, but where L1-to-L2 translation relies more strongly on word-to-word links between L1 and L2 representations than L2-to-L1 translation does, and therefore is more sensitive to cognateness. Further evidence for the presence of this difference between L1-to-L2 and L2-to-L1 translations on the degree of reliance on the lexical and or the conceptual routes between L1-to-L2 and L2-to-L1 translation was provided by @phillips2006erp, who used eletrophysiological recordings of bilinguals translating words from L1 to L2 or *vice versa* after semantic or phonological priming. The authors reported stronger N400 responses (associated with semantic priming) in L1-to-L2 translation than in L2-to-L12 translation, and stronger phonological mismatch negativity responses (associated with phonological priming) during L2-to-L1 translation, compared to L1-to-L2 translation.

@@

The existence of phonological links between words is supported with experimental evidence from the visual world paradigm and phonological priming. @meyer2007activation showed that target pictures were named faster when presented together with a phonologically-related distractor, suggesting that the activation of shared phonological information allowed the target to be activated and produced faster. There are also findings suggesting that facilitatory or inhibitory effects on target recognition depend on the degree of similarity the competitor shares with the target [@hamburger1996phonological; @dufour2003lexical]. When there is only mismatch of one phoneme between prime and target, inhibitory effects are exerted on target recognition [@dufour2003lexical]. This inhibition was proposed to be due to strong lexical activation of the prime which results in lexical interference, as opposed to simple phonological activation from less similar primes that triggers facilitation.

@@

Participants were presented with spoken words in an unfamiliar (non-native language) and were asked to translate them to their native language. We will henceforth refer to the auditory-presented words heard by participants on each trial as presented words, and the correct translation for the presented words as target translations. The motivation behind testing participants in an unfamiliar language was that these participants can be considered a particular (extreme) case of unbalanced bilingualism, in which they do not know any words in L2, and therefore word-to-concept connections are, in principle, absent. As a consequence, these participants should only be able use word-to-word connections to succeed in the translation of words in an unfamiliar language. We predicted that, if the information provided by the phonological form of the unfamiliar presented words is sufficient for translation, participants' performance should increase when the translation pairs are phonologically more similar.

Following @luce1998recognizing's NAM model, we further hypothesised that if participants are able to correctly translate words from an unfamiliar language based on their phonological similarity to the target word, similar sounding but incorrect translations (e.g., false friends) should be activated too. If this is the case, non-native words that have more phonological neighbours should be less likely to be correctly translated, especially if such neighbours have higher lexical frequency and are phonologically closer to the presented word than the target translation. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as bottom instead of as its correct translation *button* (because the lexical frequency of bottom is higher than that of button and also phonological closer). This combined constraint led us to calculate a specific form of phonological neighbours where a neighbour is counted only if it is higher frequency and has stronger phonological similarity than the target translation as judged by Levenshtein distance. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that non-native words trigger competition between words in the native language during translation.

We tested three groups of participants in this translation elicitation task: English natives translating Spanish words, English natives translating Catalan words, and Spanish natives translating Catalan words. Catalan and Spanish are both Romance languages whose close typological distance is reflected in the fact that they share many cognates, whereas English is a Germanic language that shares considerably fewer cognates with Catalan and Spanish. By testing participants translating words from typologically close and distant languages, we expected to widen the range of the phonological similarity scores of the translation pairs involved in the experimental task, therefore allowing us to explore potential cross-language differences in participant's performance.

@@

```{r}
#| label: participants
tbl_data <- participants |> 
    mutate(exp = case_when(
        source=="Experiment" & group!= "cat-SPA" ~ "Experiment 1",
        source=="Experiment" & group== "cat-SPA" ~ "Experiment 2",
        source=="Questionnaire" ~ "Experiment 3",
    )) |> 
    summarise(n = n(),
              n_excluded = sum(!valid_participant),
              across(age, lst(mean, sd, min, max)),
              knows_l2 = sum(l_2 != "None"),
              l2_lst = lst(unique(l_2)[unique(l_2)!="None"]),
              .by = c(group, exp))

@@

```{r}
#| label: participants-excluded-1
excluded_lang <- participants_1 |> 
    filter(l_2_writ_prod > 3 |
               l_2_oral_comp > 3 |
               cat_oral_comp > 3 | 
               cat_writ_prod > 3 |
               spa_oral_comp > 3 |
               spa_writ_prod > 3) |> 
    nrow()

excluded_impairment <- participants_1 |> 
    filter(has_language_problems) |> 
    nrow()

excluded_ntrials <- participants_1 |> 
    filter(invalid_participant_trials) |> 
    nrow()
```
Participants were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits), and gave informed consent before providing any data and the study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was by the University of Oxford Medical Sciences Inter-Divisional Research Ethics Committee (IDREC) (R60939/RE007). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection. We excluded data from participants that (a) self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), (b) were diagnosed with a language (*n* = `r excluded_impairment`), or (c) did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`).

@@

Each trial started with a yellow fixation point presented during one second on the centre of the screen over a black background. After one second, the audio started playing while the dot remained being displayed until the audio ended. Upon the offset of the fixation point and audio, participants were prompted to write their answer by a ">" symbol. Typed letters were displayed in the screen in real time to provide visual feed-back to participants. Participants were allowed to correct their answer. Then, participants pressed the RETURN/ENTER key to confirm their answer and start and new trial. We excluded trials where the participant's answer was not a valid word in the target language or no answer was given. Trials in which the response was mistyped by only one character were counted as correct (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct word. Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r nrow(exp_responses)` valid trials (`r nrow(exp_responses[exp_responses$group %in% c("cat-ENG", "cat-SPA"),])` in Catalan, `r nrow(exp_responses[exp_responses$group %in% c("spa-ENG"),])` in Spanish). The task took approximately 15 minutes to be completed.

```{r}
#| label: procedure-figure
#| echo: false
#| message: false
#| warning: false
#| out-width: "80%"
#| fig-cap: "Schematic representation of a trial in the experimental task."
include_graphics(here("img", "design.png"))
```

@@

We measured the phonological similarity between presented words and their target translation by computing the Levenshtein similarity between their X-SAMPA transcriptions using the `stringsim` function of the `stringdist` R package [@van2014stringdist]. This function computes the inverse of the Levenshtein distance between two character strings as a proportion. First, it computes the edit distance between two character strings (in this case, phoneme transcriptions) by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. This measure is then divided by the maximum distance (according to the length of the longest string) and then subtracted from 1. The result is a percentage score, with 0% indicating no similarity between the two strings, and 100% indicating a perfect match between the two strings. We computed this similarity measure (referred simply as Levenshtein, from here on) for every translation pair in our stimuli lists. For example, the *table* (teɪbəl)-*mesa* (mesa) translation pair had a `r percent(stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (trEIn)-*tren* (t`r knitr::asis_output("\U027E")`En) translation pair had a `r percent(stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. Figure 1 summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

We modelled the probability of participants guessing the correct target translation of each presented word using a generalised multilevel Bayesian regression model with a Bernoulli logit link distribution. We first fitted a base model that only included the intercept as a fixed effect, and random intercepts per participant  (Model 0). Secondly, we added the fixed effect of lexical frequency of the target translation ($Frequency$). We then extended the model sequentially by adding the main effect of higher-frequency phonological neighbourhood ($PTHN$), the main effect of Levenshtein similarity ($Levenshtein$), the $PTHN \times Levenshtein$ interaction, the group of participants ($Group$, with levels cat-ENG, spa-ENG, and cat-SPA), and finally, the $Levenshtein \times Group$ interaction). We specified two sum-coded *a priori* contrasts for the $Group$ predictor: one comparing English natives (cat-ENG = -0.25, spa-ENG = -0.25) against Spanish natives (cat-SPA = +0.5), and one comparing English natives translating Spanish words (spa-ENG = -0.5) against English natives translating Catalan words (cat-ENG = +0.5) [@schad2020capitalize]. Each model included random intercepts by item (to account for the correlation of responses from different participants to the same presented word), and random intercepts and slopes by participant, where our experimental designed allowed it (to account for the correlation between responses from the same participant to different presented words) [@barr2013random]. We will later refer to these models as Model 0 to Model 6. For interpretability, all continuous predictor variables were standardised (transformed in standard deviations from the mean) before entering the model.

@@

To test and account for cross-group differences, we included a random intercept for each group. We compared models using leave-one-out cross-validation (*LOO*) [@vehtari2017practical]. More information about the models and model comparison can be found in Appendix 2. All analyses were performed in R environment [@rcore2019r]. We used the tidyverse family of R packages [@wickham2019tidyverse] to process data and to generate figures. We used the brms R package [@burkner2017brms] using the cmdstanr backend to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for mode details on the models).

## Results

@@

All models showed good out-of-sample predictive validity, as suggested by the fact that the expected log-probability density was many times larger than its associated standard error. Model 6, which included all main effects and the two-way interactions between $PTHN$ and $Levenshtein$, and between $Group$ and $Levenshtein$ and $PTHN$ and consonant similarity, showed the best performance.

```{r}
#| label: model-comparison
#| message: false
#| warning: false
tab_loos <- model_loos %>% 
    as.data.frame() %>% 
    rownames_to_column("model") %>% 
    arrange(desc(elpd_loo)) %>% 
    select(model, matches("elpd_loo"), matches("looic"), matches("diff")) %>%
    mutate(model = str_replace(model, "fit_", "Model ")) %>% 
    mutate_all(~ifelse(. == 0, NA, .)) %>% 
    gt(rowname_col = "model", id = "model-comparison") %>% 
    tab_header(
        title = md("Table 2"),
        subtitle = md("Model performance and comparison using leave-one-out cross-validation")
    ) %>% 
    tab_source_note(
        md("*Note*. Outcomes of the leave-one-out cross-validation procedure. Each row indicates the values of the expected log-predicted density (*ELPD*), leave-one-out information creterion (*LOO-IC*), and the difference in *ELPD* (*LOO-diff*) between each model and the inmediately simpler model. Each value is accompanied by a standard error (*SE*) indicating the uncertainty of its estimate.")
    ) %>% 
    fmt_missing(everything(), missing_text = "-") %>% 
    fmt_number(2:7) %>% 
    cols_label(
        model = md("**Model**"),
        looic = md("***LOO<sub>IC</sub>***"),
        se_looic = md("***SE*<sub>IC</sub>**"),
        elpd_loo = md("***LOO<sub>ELPD</sub>***"),
        se_elpd_loo = md("***SE***"),
        elpd_diff = md("***LOO<sub>diff</sub>***"),
        se_diff = md("***SE<sub>diff</sub>***")
    ) %>% 
    tab_style(
        style = cell_text(align = "left", weight = "bold", size = "medium"),
        locations = cells_title(groups = "title")
    ) %>% 
    tab_style(
        style = cell_text(align = "left", size = "medium", style = "italic"),
        locations = cells_title(groups = "subtitle")
    ) %>% 
    tab_style(
        style = cell_text(align = "left", size = "medium"),
        locations = cells_source_notes()
    ) %>% 
    tab_style(
        style = list(
            cell_borders(sides = c("left", "right"), color = "white"),
            cell_text(align = "center")
        ),
        locations = list(
            cells_body(), 
            cells_row_groups(), 
            cells_column_spanners(), 
            cells_stub(), 
            cells_source_notes(),
            cells_column_labels(), 
            cells_stubhead()
        )
    ) %>% 
    tab_style(
        style = list(
            cell_text(align = "left", size = "medium"), 
            cell_borders(sides = "all", color = "white")
        ),
        locations = list(cells_stub(), cells_source_notes())
    ) %>% 
    tab_style(
        style = cell_text(align = "center", size = "medium", weight = "normal", style = "normal"),
        locations = list(cells_column_spanners(), cells_column_labels())
    ) %>% 
    tab_style(
        style = cell_borders(sides = "all", color = "white"),
        locations = cells_body(columns = 1:7)
    ) %>% 
    tab_style(
        style = cell_borders(sides = "top", color = "white"),
        locations = cells_body(columns = 1:7)
    ) %>% 
    gtsave(here("img", "loos.png"))

include_graphics(here("img", "loos.png"))

@@

# Discussion

>>>>

library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(gt)
library(gtExtras)
library(brms)
library(tidybayes)

@@

# import objects
tar_config_set(store = here::here("_targets"),
               script = here::here("_targets.R"))
tar_load_globals()
tar_load(participants)
tar_load(exp_participants)
tar_load(quest_participants)
tar_load(stimuli)
tar_load(epreds)

theme_set(theme_ggdist() + theme(panel.grid.major.y = element_line(colour = "grey",
                                                                   linetype = "dotted")))

@@

Listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. However, tfile.path are some sources of information which may make non-native speech comprehension easier. One of the sources of this increased difficulty stems from the possible mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener’s native language. For example, imagine a Spanish native, with no previous familiarity with any other language, who listens for the first time to French may encounter the word /`r knitr::asis_output("\U0070\U0254\U0281\U0074")`/ (*porte*), translation of door in French. The voiced uvular fricative consonant /`r knitr::asis_output("\U0281")`/ and the open-mid back rounded vowel /`r knitr::asis_output("\U0254")`/ do not exist in Spanish as phonemic categories. These discrepancies can nonetheless be perceived as allophonic variations the native (Spanish) phonemes /r/ and /o/ [@best2001discrimination], and the non-native speech signal may engage lexical processing mechanisms [@weber2004lexical], leading to the activation and selection, and tfile.pathfore the recognition of its translation to Spanish /`r knitr::asis_output("\U02C8\U0070\U0077\U0065\U0072\U002E\U0074\U0061")`/ (*puerta*). The recognition of this French, unfamiliar word relies almost entirely on its similarity with its Spanish translation. Conversely, imagine now that the same Spanish listener hears English for the first time and encounters the word /`r knitr::asis_output("\U0064\U0254\U02D0")`/ (*door*). This time, phonological similarity is of little help: /`r knitr::asis_output("\U02C8\U0070\U0077\U0065\U0072\U002E\U0074\U0061")`/ and /`r knitr::asis_output("\U0064\U0254\U02D0")`/ share very few phonemes. It would come as a surprise if the listener was able to translate the English word successfully, if not by relying on previously learnt word-concept associations (which they lack, as they are unfamiliar to English).

@@

Early theoretical accounts of this 'Cognate effect' proposed that the impact of cognateness on lexical processing depends on bilinguals’ proficiency in their second language (L2) [@potter1984lexical; see @chen1989patterns for similar results]. The reasoning behind this claim was the assumption that second-language learners start acquiring words in their L2 by associating them to their translation in their L1 (lexical route), instead of directly to their shared concept (conceptual route). This word-to-word connectivity should be sensitive to form-similarity (i.e., cognateness): the more similar the two word- forms are, the stronger the connection. As learners become more proficient, the connection between L2 representations and their meaning grows stronger, and L2 word processing becomes less reliant on word-word connections between L1 and L2 representations. Cognates subsequently exert less impact on lexical processing in proficient speakers [see @andras2022cognate for recent experimental evidence for this claim]. The Revised Hierarchical Model [RHM, @kroll1994category] captured this hypothesis and predicted that translating words from L1 to L2 should take longer than translating from L2 to L1. The rationale behind this prediction was that L1-to-L2 translation relies more strongly on word-to-word links between L1 and L2 representations, wfile.pathas L2-to-L1 translation relies more strongly on the mediation between the concept and the two word- forms. L1-to-L2 translation should tfile.pathfore be more sensitive to the form similarity between the L1 and the L2 representations: cognate words should be retrieved faster than non-cognate words during L1-to-L2 translation.

@@

These findings support a soft version of the RHM model, in which both the conceptual and the lexical translation routes are active during translation, but wfile.path L1-to-L2 translation relies more strongly on word-to-word links between L1 and L2 representations than L2-to-L1 translation does, and tfile.pathfore is more sensitive to cognateness. Further evidence for the presence of this difference between L1-to-L2 and L2-to-L1 translations on the degree of reliance on the lexical and or the conceptual routes between L1-to-L2 and L2-to-L1 translation was provided by @phillips2006erp, who used eletrophysiological recordings of bilinguals translating words from L1 to L2 or *vice versa* after semantic or phonological priming. The authors reported stronger N400 responses (associated with semantic priming) in L1-to-L2 translation than in L2-to-L12 translation, and stronger phonological mismatch negativity responses (associated with phonological priming) during L2-to-L1 translation, compared to L1-to-L2 translation.

@@

The existence of phonological links between words is supported with experimental evidence from the visual world paradigm and phonological priming. @meyer2007activation showed that target pictures were named faster when presented together with a phonologically-related distractor, suggesting that the activation of shared phonological information allowed the target to be activated and produced faster. Tfile.path are also findings suggesting that facilitatory or inhibitory effects on target recognition depend on the degree of similarity the competitor shares with the target [@hamburger1996phonological; @dufour2003lexical]. When tfile.path is only mismatch of one phoneme between prime and target, inhibitory effects are exerted on target recognition [@dufour2003lexical]. This inhibition was proposed to be due to strong lexical activation of the prime which results in lexical interference, as opposed to simple phonological activation from less similar primes that triggers facilitation.

@@

Participants were presented with spoken words in an unfamiliar (non-native language) and were asked to translate them to their native language. We will henceforth refer to the auditory-presented words heard by participants on each trial as presented words, and the correct translation for the presented words as target translations. The motivation behind testing participants in an unfamiliar language was that these participants can be considered a particular (extreme) case of unbalanced bilingualism, in which they do not know any words in L2, and tfile.pathfore word-to-concept connections are, in principle, absent. As a consequence, these participants should only be able use word-to-word connections to succeed in the translation of words in an unfamiliar language. We predicted that, if the information provided by the phonological form of the unfamiliar presented words is sufficient for translation, participants' performance should increase when the translation pairs are phonologically more similar.

Following @luce1998recognizing's NAM model, we further hypothesised that if participants are able to correctly translate words from an unfamiliar language based on their phonological similarity to the target word, similar sounding but incorrect translations (e.g., false friends) should be activated too. If this is the case, non-native words that have more phonological neighbours should be less likely to be correctly translated, especially if such neighbours have higher lexical frequency and are phonologically closer to the presented word than the target translation. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as bottom instead of as its correct translation *button* (because the lexical frequency of bottom is higher than that of button and also phonological closer). This combined constraint led us to calculate a specific form of phonological neighbours wfile.path a neighbour is counted only if it is higher frequency and has stronger phonological similarity than the target translation as judged by Levenshtein distance. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that non-native words trigger competition between words in the native language during translation.

We tested three groups of participants in this translation elicitation task: English natives translating Spanish words, English natives translating Catalan words, and Spanish natives translating Catalan words. Catalan and Spanish are both Romance languages whose close typological distance is reflected in the fact that they share many cognates, wfile.pathas English is a Germanic language that shares considerably fewer cognates with Catalan and Spanish. By testing participants translating words from typologically close and distant languages, we expected to widen the range of the phonological similarity scores of the translation pairs involved in the experimental task, tfile.pathfore allowing us to explore potential cross-language differences in participant's performance.

@@

```{r}
#| label: participants
tbl_data <- participants |> 
    mutate(exp = case_when(
        source=="Experiment" & group!= "cat-SPA" ~ "Experiment 1",
        source=="Experiment" & group== "cat-SPA" ~ "Experiment 2",
        source=="Questionnaire" ~ "Experiment 3",
    )) |> 
    summarise(n = n(),
              n_excluded = sum(!valid_status=="Valid"),
              across(age, lst(mean, sd, min, max)),
              knows_l2 = sum(l2 != "None"),
              l2_lst = lst(unique(l2)[unique(l2)!="None"]),
              .by = c(group, exp))

@@

```{r}
#| label: participants-excluded-1
excluded_lang <- participants_1 |> 
    filter(l2_writ_prod > 3 |
               l2_oral_comp > 3 |
               cat_oral_comp > 3 | 
               cat_writ_prod > 3 |
               spa_oral_comp > 3 |
               spa_writ_prod > 3) |> 
    nrow()

excluded_impairment <- participants_1 |> 
    filter(valid_status=="Language impairment") |> 
    nrow()

excluded_ntrials <- participants_1 |> 
    filter(valid_status=="Insufficient trials") |> 
    nrow()
```
Participants were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits), and gave informed consent before providing any data and the study was conducted in accordance with ethical standards of the Declaration of Helsinki and the protocol was by the University of Oxford Medical Sciences Inter-Divisional Research Ethics Committee (IDREC) (R60939/RE007). Participants were asked to complete the experiment using a laptop in a quiet place with good internet connection. We excluded data from participants that (a) self-rated their oral and/or written skills in Catalan and Spanish, or any other second language as four or higher in a five-point scale (*n* = `r excluded_lang`), (b) were diagnosed with a language (*n* = `r excluded_impairment`), or (c) did not contribute more than 80% of valid trials (*n* = `r excluded_ntrials`).

@@

Each trial started with a yellow fixation point presented during one second on the centre of the screen over a black background. After one second, the audio started playing while the dot remained being displayed until the audio ended. Upon the offset of the fixation point and audio, participants were prompted to write their answer by a ">" symbol. Typed letters were displayed in the screen in real time to provide visual feed-back to participants. Participants were allowed to correct their answer. Then, participants pressed the RETURN/ENTER key to confirm their answer and start and new trial. We excluded trials wfile.path the participant's answer was not a valid word in the target language or no answer was given. Trials in which the response was mistyped by only one character were counted as correct (e.g., "pengiun" instead of "penguin"), as long as the response did not correspond to a distinct word. Trials in which participants took longer than 10 seconds to respond were also excluded. Participants contributed a total of `r nrow(exp_responses)` valid trials (`r nrow(exp_responses[exp_responses$group %in% c("cat-ENG", "cat-SPA"),])` in Catalan, `r nrow(exp_responses[exp_responses$group %in% c("spa-ENG"),])` in Spanish). The task took approximately 15 minutes to be completed.

```{r}
#| label: procedure-figure
#| echo: false
#| message: false
#| warning: false
#| out-width: "80%"
#| fig-cap: "Schematic representation of a trial in the experimental task."
knitr::include_graphics(here::here("img", "design.png"))
```

@@

We measured the phonological similarity between presented words and their target translation by computing the Levenshtein similarity between their X-SAMPA transcriptions using the `stringsim` function of the `stringdist` R package [@van2014stringdist]. This function computes the inverse of the Levenshtein distance between two character strings as a proportion. First, it computes the edit distance between two character strings (in this case, phoneme transcriptions) by counting the number of additions, deletions, and substitutions necessary to make both strings identical [@levenshtein1966binary]. This measure is then divided by the maximum distance (according to the length of the longest string) and then subtracted from 1. The result is a percentage score, with 0% indicating no similarity between the two strings, and 100% indicating a perfect match between the two strings. We computed this similarity measure (referred simply as Levenshtein, from file.path on) for every translation pair in our stimuli lists. For example, the *table* (teɪbəl)-*mesa* (mesa) translation pair had a `r percent(stringsim(enc2utf8("teɪbəl"), enc2utf8("mesa")))` similarity, while the *train* (trEIn)-*tren* (t`r knitr::asis_output("\U027E")`En) translation pair had a `r percent(stringsim(enc2utf8("trEIn"), enc2utf8("tɾEn")))` similarity. Figure 1 summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

We modelled the probability of participants guessing the correct target translation of each presented word using a generalised multilevel Bayesian regression model with a Bernoulli logit link distribution. We first fitted a base model that only included the intercept as a fixed effect, and random intercepts per participant  (Model 0). Secondly, we added the fixed effect of lexical frequency of the target translation ($Frequency$). We then extended the model sequentially by adding the main effect of higher-frequency phonological neighbourhood ($PTHN$), the main effect of Levenshtein similarity ($Levenshtein$), the $PTHN \times Levenshtein$ interaction, the group of participants ($Group$, with levels cat-ENG, spa-ENG, and cat-SPA), and finally, the $Levenshtein \times Group$ interaction). We specified two sum-coded *a priori* contrasts for the $Group$ predictor: one comparing English natives (cat-ENG = -0.25, spa-ENG = -0.25) against Spanish natives (cat-SPA = +0.5), and one comparing English natives translating Spanish words (spa-ENG = -0.5) against English natives translating Catalan words (cat-ENG = +0.5) [@schad2020capitalize]. Each model included random intercepts by item (to account for the correlation of responses from different participants to the same presented word), and random intercepts and slopes by participant, wfile.path our experimental designed allowed it (to account for the correlation between responses from the same participant to different presented words) [@barr2013random]. We will later refer to these models as Model 0 to Model 6. For interpretability, all continuous predictor variables were standardised (transformed in standard deviations from the mean) before entering the model.

@@

To test and account for cross-group differences, we included a random intercept for each group. We compared models using leave-one-out cross-validation (*LOO*) [@vehtari2017practical]. More information about the models and model comparison can be found in Appendix 2. All analyses were performed in R environment [@rcore2019r]. We used the tidyverse family of R packages [@wickham2019tidyverse] to process data and to generate figures. We used the brms R package [@burkner2017brms] using the `cmdstanr` backend to the Stan probabilistic language [@carpenter2017stan] to estimate and compare the models (see Appendix 1 for mode details on the models).

### Results

@@

```{r exp_1_epreds}
#| label: fig-epreds-1
#| fig-height: 3.5
#| fig-width: 10
get_epreds(exp_1_m0, dataset_1) |> 
    mutate(neigh_n_h = paste0(neigh_n_h, " neighbours"),
           neigh_n_h = factor(neigh_n_h,
                              levels = paste0(c(0, 2, 4, 8, 16),
                                              " neighbours"),
                              ordered = TRUE)) |> 
    ggplot(aes(lv, .epred)) +
    facet_wrap(~neigh_n_h, nrow = 1) +
    stat_lineribbon(aes(fill_ramp = after_stat(level)),
                    linewidth = 1/2,
                    .width = c(0.99, 0.95, 0.89, 0.78, 0.67, 0.50)) +
    labs(x = "Cognateness\n(Levenshtein distance with correct translation)",
         y = "p(Correct)",
         colour = "Cross-language neighbourhood density",
         fill = "Cross-language neighbourhood density",
         linetype = "Cross-language neighbourhood density",
         fill_ramp = "CrI (%)") +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_brewer(palette = "Reds") +
    theme(legend.position = "none",
          legend.box = "vertical") 
```

@@

### Discussion

## Experiment 2

### Methods

### Results

```{r exp_2_epreds}
#| label: fig-epreds-2
#| fig-height: 3.5
#| fig-width: 10
get_epreds(exp_2_m0, dataset_2) |> 
    mutate(neigh_n_h = paste0(neigh_n_h, " neighbours"),
           neigh_n_h = factor(neigh_n_h,
                              levels = paste0(c(0 , 1, 2, 4, 8, 16),
                                              " neighbours"),
                              ordered = TRUE)) |>
    ggplot(aes(lv, .epred)) +
    facet_wrap(~neigh_n_h, nrow = 1) +
    stat_lineribbon(aes(fill_ramp = after_stat(level)),
                    linewidth = 1/2,
                    .width = c(0.99, 0.95, 0.89, 0.78, 0.67, 0.50)) +
    labs(x = "Cognateness\n(Levenshtein distance with correct translation)",
         y = "P(Correct)",
         colour = "Cross-language neighbourhood density",
         fill = "Cross-language neighbourhood density",
         linetype = "Cross-language neighbourhood density",
         fill_ramp = "CrI (%)") +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_brewer(palette = "Reds") +
    theme(legend.position = "none",
          legend.box = "vertical") 
```

### Discussion

## Experiment 3

```{r exp_3_epreds}
#| label: fig-epreds-3
#| fig-height: 3.5
#| fig-width: 10
get_epreds(exp_3_m0, dataset_3) |> 
    mutate(neigh_n_h = paste0(neigh_n_h, " neighbours"),
           neigh_n_h = factor(neigh_n_h,
                              levels = paste0(c(0 , 1, 2, 4, 8, 16),
                                              " neighbours"),
                              ordered = TRUE)) |> 
    ggplot(aes(lv, .epred)) +
    facet_wrap(~neigh_n_h, nrow = 1) +
    stat_lineribbon(aes(fill_ramp = after_stat(level)),
                    linewidth = 1/2,
                    .width = c(0.99, 0.95, 0.89, 0.78, 0.67, 0.50)) +
    labs(x = "Cognateness\n(Levenshtein distance with correct translation)",
         y = "p(Correct)",
         colour = "Cross-language neighbourhood density",
         fill = "Cross-language neighbourhood density",
         linetype = "Cross-language neighbourhood density",
         fill_ramp = "CrI (%)") +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_brewer(palette = "Reds") +
    theme(legend.position = "none",
          legend.box = "vertical") 
```

### Methods

### Results

### Discussion
